Bootstrap learning via modular concept discovery

### 2013-01-29 drafts/ijcai2013/ijcai13.tex
Changes.

Formally then, our goal is to solve the following minimal feedback
multi-task program induction problem: suppose we are
given set of \emph{tasks} $T=\{t_k\}_{k=1}^K$ where each task is a
function $t_i : \mathcal{L} \rightarrow \{0, 1\}$, where $\mathcal{L}$
is the set of programs in our language.  We say that $ p \in
\mathcal{L}$ \emph{solves} $t_i$ if $ t_i\; p = 1$. Our
goal is to find a set of solutions $X = \{x_k\}_{k=1}^k \subset
\mathcal{L}$ such that

\[
X = \argmax_{x_1, \dots, x_K \in L} \sum_k t_k(x_k).
\]

Based on the intuition set out above, we present an iterative
strategy. Given a stochastic grammar $\mathcal{G}$ which defines a
distribution over the primitives from $\mathcal{L}$ and a integer
\emph{frontier size} $N$, we perform to steps in each iteration.  In
the first step, we explore the $N$ highest probability programs
according $\mathcal{G}$. In the second step, we compress the
programs that solve any of our tasks, and we use that compression to
define a new grammar. Repeating these two steps, we effectively
reparameterize our search space to include programs that are more likely
to be solutions to our tasks.

@@

\subsection{Representing typed functional programs as binary trees.}

@@

The basis combinators can themselves been expressed in the lambda
calculus. The lambda calculus has two basic operation -- application
and abstraction -- but in using the combinatory logic we sequester
uses of the abstraction operation inside the combinators, and thus
have a representation language that is effectively abstraction- and,
thus, variable-free. In doing so, we have replaced the variable
binding of the $\lambda$ operator with the variable \emph{routing} of
these basis combinators. See~\cite{liang10programs} for a more
detailed discussion of this routing interpretation.

@@

We can extend this representation with a simple polymorphic type
system~\cite{Pierce_2002}. Thus, this combinator can be represented as
a binary tree~(Figure~\ref{fig:clbintree}) whose leaves are typed
primitive combinators and whose interior nodes represent typed applications
of combinators.

@@

\section{Experiments.}
\subsection{Symbolic regression.}

@@

\subsection{Boolean circuits}

As another demonstration of this approaches learning abilities
investigate its ability to learn boolean functions using only the
logical primitive NAND and the basic combinators. In
Figure~\ref{fig:boolLearningCurves},

\begin{figure}
\begin{subfigure}[Before]{0.45\linewidth}
\includegraphics[width=\linewidth]{figures/boolLearningCurves.pdf}
\end{subfigure}
\caption{(a) Average learning curves as we vary the frontier size. (b)
  Percent of tasks solved at iteration 15 as a function of frontier
  size.}
\label{fig:boolLearningCurves} 
\end{figure}

@@

\begin{figure}
\begin{circuitikz} \draw
%%(1,0) node[nand port, thick] (nor1) {}
(1,.75) node[nand port, thick] (nor1) {}
(3,0) node[nand port, thick] (nor2) {}
(3,1.5) node[nand port, thick] (nor3) {}
(5,.75) node[nand port, thick] (nor4) {}

(nor1.in 1) node[above left=.5cm] (x) {x}
(nor1.in 2) node[below left = .5cm](y) {y}
(x) -| (nor1.in 1)
(y) -| (nor1.in 2)
(x) -| (nor3.in 1)
(y) -| (nor2.in 2)
(nor1) -- (1.62, .75) -- (nor2.in 1)
(nor1) -- (1.62, .75) -- (nor3.in 2)
(nor2) -| (nor4.in 2)
(nor3) -| (nor4.in 1);
 \end{circuitikz}
\caption{A standard XOR circuit built of NAND gates.}
\label{fig:xor_circuit}
\end{figure}

\begin{figure}
\begin{circuitikz} \draw

%%(1,0) node[nand port, thick] (nor1) {}
(3, 0)  node[ anchor=east, text align=center, thick, draw, text width=1cm, text height=.25cm, align=center] (f) {?} 
node [right of=f] (out) {}
(1, 0) node[nand port, thick] (nor) {}
(nor.in 1) node[above left=.25cm] (x) {x}
(nor.in 2) node[below left=.25cm] (y) {y}
(y) -| (f)
(nor) -- (f)
(f) -| (out)
(x) -| (nor.in 1) 
(y) -| (nor.in 2);
\end{circuitikz}
\caption{A schematic that represents the reused component
  \texttt{\small{(B (C B nand) S)}}.}
\label{fig:component}
\end{figure}

@@

\subsection{Discussion}
...

>>>>

Let a \emph{task} be a function $t : \mathcal{L} \rightarrow \{0,
1\}$, where $\mathcal{L}$ is the set of all expressions in our
representation language. We say that $e \in \mathcal{L}$ \emph{solves}
task $t$ if $t(e) = 1$. Our goal is to solve as many tasks as
possible, as efficiently as possible. That is, suppose that there is a
distribution over tasks $t \sim  p_t(\cdot)$, we would like an ordering $O
= e_1, \dots $ on expressions in $\mathcal{L}$ such that if $p_t(t_1)
> p_t(t_2)$ then the first expression in $O$ to solve $t_1$ comes
before the first expression in $O$ to solve $t_2$.

We motivate a solution to this problem by considering a heirarchical
Bayesian formulation of the task distribution. Suppose we believe that
there is some underlying structure to $p_t(\cdot)$ that is explained by
the fact that for each task $t$ there is a latent expressions $e \in
\mathcal{L}$ drawn from a stochastic grammar $\mathcal{G}$ and that

\begin{align}
e &\sim p_\mathcal{L}(\cdot) = \mathcal{G}\\
t | e &\sim \text{U}\{ \text{tasks solved by }e\}.
\end{align}

Then $p_t(t) = \sum_{e_j \in \mathcal{L} \text{ s.t. } t(e_j) = 1}
p_\mathcal{L}(e_j)$. If we make the assumption that for the set $E$ of
expressions in this sum, there is some expression $e^*$ such that
$p_\mathcal{L}(e^*) >> p_\mathcal{L}(e), e \neq e^*$, then we have, approximately, 
\[ p_t(t) \propto} p_\mathcal{L}(e^*). \]

This suggests a way to find the ordering $O$: infer a stochastic
grammar $G$, under the assumption that $G$ favors one solution to a
given task over any other, and order the elements of $\mathcal{L}$
according to $G$.

Thus, our basic procedure will be to search for a single ``good''
solution for as many tasks as we can, and to use those solutions to
estimate a stochastic grammar. Our search will be performed in the
order of the most recently inferred grammar. So as our ordering
becomes more efficient and we can solve more tasks, we learn better
stochastic grammars.

%% Formally then, our goal is to solve the following minimal feedback
%% multi-task program induction problem: suppose we are given set of
%% \emph{tasks} $T=\{t_k\}_{k=1}^K$ where each task is a function $t_i :
%% \mathcal{L} \rightarrow \{0, 1\}$, where $\mathcal{L}$ is the set of
%% programs in our language.  We say that $ p \in \mathcal{L}$
%% \emph{solves} $t_i$ if $ t_i\; p = 1$. Our goal is to find a set of
%% solutions $X = \{x_k\}_{k=1}^k \subset \mathcal{L}$ such that

%% \[
%% X = \argmax_{x_1, \dots, x_K \in L} \sum_k t_k(x_k).
%% \]

Based on the motivation set out above, we present an iterative
strategy. Given a stochastic grammar $\mathcal{G}$ which defines a
distribution over the primitives from $\mathcal{L}$ and a integer
\emph{frontier size} $N$, we perform to steps in each iteration.  In
the first step, we explore the $N$ highest probability programs
according to $\mathcal{G}$. In the second step, we compress the
programs that solve any of our tasks, and we use that compression to
define a new grammar. Repeating these two steps, we effectively
reparameterize our search space to include programs that are more likely
to be solutions to our tasks.

@@

\subsection{Representing typed functional programs as binary trees}

@@

The basis combinators can themselves been expressed in the lambda
calculus. The lambda calculus has two basic operations -- application
and abstraction -- but in using the combinatory logic we sequester
uses of the abstraction operation inside the combinators, and thus
have a representation language that is effectively abstraction- and,
thus, variable-free. In doing so, we have replaced the variable
binding of the $\lambda$ operator with the variable \emph{routing} of
these basis combinators. See~\cite{liang10programs} for a more
detailed discussion of this routing interpretation.

@@

We can extend this representation with a simple polymorphic type
system~\cite{Pierce_2002}. In this representation, a type is either a
type primitive (e.g. reals, integers, booleans, etc.), a type variable
(e.g. $\sigma$), or a function type $ t_1 \rightarrow t_2$ of
functions from source type $t_1$ to target type $t_2$. Thus, this
combinator can be represented as a binary
tree~(Figure~\ref{fig:clbintree}) whose leaves are typed primitive
combinators and whose interior nodes represent typed applications of
combinators.

@@

\section{Experiments.}
\subsection{Symbolic regression.}
\label{sec:symreg}

@@

\subsection{Boolean circuits}
\label{sec:boolcircuits}

As another demonstration of this approach, we investigate the
procedure's ability to learn boolean functions using only the logical
primitive NAND and the basic combinators. It is well known that a
boolean circuit can be constructed for any boolean function given only
the NAND gate. Here, the basic combinators take the place of the
wiring normally found in a circuit, the output of element to multiple
inputs downstream.

We constructed two task sets. The first of these was constructed
explicitly to contain familiar module structure. In this set of tasks,
we sampled $1000$ boolean circuits using AND, OR, and NOT gates. To
accomplish this, we first sampled either 1, 2, 3 inputs, then between
1 and 5 gates, randomly wiring the inputs of each new gate to the
output of one of the existing gates in the circuit. We continued this
sampling procedure until we had 1000 ``connected'' circuits, i.e.,
circuits all of whose output are wired to an input except for the last
one (the output gate). This resulted in $1000$ tasks consisting of
$82$ unique boolean functions, with a distribution of tasks as shown
in Figure~\ref{fig:boolhist}. In Figure~\ref{fig:boolLearningCurves1},
We present learning curves for frontier sizes 100, 500 and 10000,
using the same procedure as in Section~\ref{sec:symreg}.

This problem is particularly suitable for inspecting the constituents
of the induced grammar $\mathcal{G}$. We might hope that our algorithm
recovers the consituent logic gates that were used to build up the
problems. Table~\ref{table:bool} shows a few of the top ranked
expressions in the library. These include the basic logic gates; we
also include one higher-order expression $F$, to stress that the
representation we are using allows one to build concepts that are more
expressive that just sub-circuits.

In a second experiment, we include all $272$ boolean truth tables with
three or fewer inputs as tasks. This is a more difficult problem. Note
that in the first boolean function problem, there are two kinds of
learning that can take place: first, many expressions map to a single
boolean function, so we need to learn primitives that allow us to span
many different functions instead of allows generating very simple
functions over and over again. The second kind of learning involves
the distribution of the functions themselves. In the first boolean
experiment, that structure is apparent in the distribution in
Figure~\ref{fig:boolhist}. In this second experiment, we remove the
second kind of structure. In Figure~\ref{boolLearningCurves}, we show
10 runs of the algorithm with a frontier size of $2000$: note how
there are several local minima that majority of the runs get stuck in
with performance around $50\%$, but several of the runs seem to take
different trajectories to more successful representations.

\begin{figure}
\begin{subfigure}[Before]{0.45\linewidth}
\includegraphics[width=.9\linewidth]{figures/booleancircuitsBars.pdf}
\end{subfigure}
\begin{subfigure}[Before]{0.45\linewidth}
\includegraphics[width=.9\linewidth]{figures/boolLearningCurves.pdf}
\end{subfigure}
\caption{(a) Average learning curves as we vary the frontier size. (b)
  Percent of tasks solved at iteration 15 as a function of frontier
  size.}
\label{fig:boolLearningCurves}

@@

\begin{table}
\scriptsize
  \begin{tabular}{|l|l|l|}
    function  & CL expressions & circuit  ~ \\ \hline
        TRUE  & (S NAND) (S NAND I)                   & ~ \\ \hline
        FALSE & (S B (S NAND)) (S NAND I)             & ~ \\ \hline
        NOT   & S NAND I                              & \input{figures/NOT.tex} \\ \hline
        AND   & ((C B NAND) (B (S NAND I)))           & ~ \\ 
        OR    & ((B (C (B (S NAND) NAND))) (S NAND I) & ~ \\ 
        XOR   & ~                                     & ~ \\ 
        F     & S B (S NAND)                          & \input{figures/F.tex}
    \end{tabular}
\caption{Expressions in the induced library on sets of tasks
  corresponding to boolean circuits synthesized from AND, OR, and NOT
  gates, with boolean circuits of 1, 2, or 3 inputs.}
\label{table:boolexpr}
\end{table}

@@

\subsection{Discussion}
